---
title: "Stage_2_Explorations"
author: "Juan Pablo"
date: "2024-09-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Stage 2

I worked in R for the data explorations just because of the ease of seeing tables and variables in the environment. I hope that's ok! I usually work in R Scripts but exporting to html is easier through Rmd, so here we are :)

```{r}
#libraries
library(tidyverse)
library(here)
library(httr)
library(jsonlite)

```


## Explorations - Voting Data

```{r}

# set directory
here::i_am("code/exploration_votingdata.R")

county_pres <- read_csv(here("data/countypres_2000-2020.csv"))

# USING FIPS ----------------------------------------------------------------


# number of counties in each year 
counties_in_year <- county_pres %>%
  group_by(year) %>%
  summarise(num_counties = n_distinct(county_fips))
# this seems too good to be true; need to investigate


# looking at unique years in each county
year_fips_combinations <- county_pres %>%
  distinct(year, county_fips) %>%
  group_by(county_fips) %>%
  mutate(years_present = n_distinct(year))
# not so helpful; still a super manual task



#go one layer down and count how many distinct years each county has
counts_years <- county_pres %>% 
  group_by(county_fips) %>%
  summarise(year_counts = n_distinct(year))
# this looks pretty good

appearances_distribution <- counts_years %>%
  group_by(year_counts) %>%
  summarise(n_counties = n_distinct(county_fips))

# counties to get rid of
counties_troublesome <- year_fips_combinations %>%
  filter(years_present < 6)


county_pres %>%
  summarise(unique_count = n_distinct(county_fips))



```


## Explorations - Census Data
Note on Census data: I have hard copy csvs for 2012, 2016, and 2020, but for some reason the earlier data is only accessible through API. Strange, but that's why we have the api here. 

```{r}



# set directory
here::i_am("code/exploration_censusdata.R")

# US Census Board API Key
# 6b04ff55736c418eed28d02b041d8a382d5b0319
USCB_APIkey <- "6b04ff55736c418eed28d02b041d8a382d5b0319"



# API FETCH -------------------------------------------------------------------
# Basic GET request to the API
#["DP02_0060PE",
#"Percent!!EDUCATIONAL ATTAINMENT!!Population 25 years and over!!9th to 12th grade, no diploma",
#"SELECTED SOCIAL CHARACTERISTICS IN THE UNITED STATES"]

# 2008 does not have 5 year data, as the survey began in 2005
# 2009 DOES have 5 year data
#url <- "https://api.census.gov/data/2008/acs/acs1/profile?get=DP02_0060PE&for=county:*&key=6b04ff55736c418eed28d02b041d8a382d5b0319"  
url <- "https://api.census.gov/data/2009/acs/acs5/profile?get=DP02_0060PE&for=county:*&key=6b04ff55736c418eed28d02b041d8a382d5b0319"  


# need to convert this into a function so i can iterate over all the columns i want for 2008
response <- GET(url) # request
data <- content(response, as = "text", encoding = "UTF-8") # data encoding for json
json_data <- fromJSON(data) #json extract
json_as_df <- as.data.frame(json_data[-1, ])  # skip first row for header
colnames(json_as_df) <- json_data[1, ]  # assign column name






# EPLORATIONS -------------------------------------------------------------
# Decennial ----------------------------------------------------------------
# load data
Decennial_2000 <- read_csv(here("data/DECENNIALDPSF42000.DP1_2024-10-19T161201/DECENNIALDPSF42000.DP1-Data.csv"))

Decennial_2000_edited <- Decennial_2000 %>%
  select(-contains("Percent"),
         -contains("Margin"))

head(names(Decennial_2000_edited), 5)
# Total population!!SEX AND AGE
# RELATIONSHIP
# HOUSEHOLDS BY TYPE
# HOUSING TENURE



# DP2 - Social ----------------------------------------------------------------
# load data
DP2_2012_5Y <- read_csv(here("data/productDownload_2024-10-19T191219/ACSDP5Y2012.DP02-Data.csv"))

DP2_2012_5Y_edited <- DP2_2012_5Y %>%
  select(-contains("Percent"),
         -contains("Margin"),
         -contains("GRANDPARENTS"))

head(names(DP2_2012_5Y_edited), 5)
# HOUSEHOLDS BY TYPE
# RELATIONSHIP
# MARITAL STATUS
# FERTILITY
# SCHOOL ENROLLMENT
# EDUCATIONAL ATTAINMENT
# VETERAN STATUS
# DISABILITY STATUS OF THE CIVILIAN NONINSTITUTIONALIZED POPULATION
# RESIDENCE 1 YEAR AGO
# PLACE OF BIRTH
# U.S. CITIZENSHIP STATUS
# YEAR OF ENTRY
# WORLD REGION OF BIRTH OF FOREIGN BORN
# LANGUAGE SPOKEN AT HOME
# ANCESTRY

# Function to count "null" values in each column
count_nulls <- function(col) {
  sum(tolower(trimws(col)) == "null", na.rm = TRUE)
}

# Apply the function to each column in the dataframe
null_counts <- sapply(DP2_2012_5Y_edited, count_nulls)
# you can check this out to see it's 78 across all






# DP3 - Economic ----------------------------------------------------------------
# load data
DP3_2012_5Y <- read_csv(here("data/productDownload_2024-10-19T191310/ACSDP5Y2012.DP03-Data.csv"))

DP3_2012_5Y_edited <- DP2_2012_5Y %>%
  select(-contains("Percent"),
         -contains("Margin"))

head(names(DP3_2012_5Y_edited), 5)
# EMPLOYMENT STATUS
# COMMUTING TO WORK
# OCCUPATION
# INDUSTRY
# CLASS OF WORKER
# INCOME AND BENEFITS (IN 2012 INFLATION-ADJUSTED DOLLARS)
# HEALTH INSURANCE COVERAGE




# DP4 - Selected Housing ----------------------------------------------------------------
# load data
DP4_2012_5Y <- read_csv(here("data/productDownload_2024-10-19T191541/ACSDP5Y2012.DP04-Data.csv"))

DP4_2012_5Y_edited <- DP4_2012_5Y %>%
  select(-contains("Percent"),
         -contains("Margin"))

head(names(DP4_2012_5Y_edited), 5)
# HOUSING OCCUPANCY
# UNITS IN STRUCTURE
# YEAR STRUCTURE BUILT
# ROOMS
# BEDROOMS
# HOUSING TENURE
# YEAR HOUSEHOLDER MOVED INTO UNIT
# VEHICLES AVAILABLE
# HOUSE HEATING FUEL
# SELECTED CHARACTERISTICS
# OCCUPANTS PER ROOM
# VALUE
# MORTGAGE STATUS
# SELECTED MONTHLY OWNER COSTS (SMOC)
# GROSS RENT





# DP4 - Selected Housing ----------------------------------------------------------------
# load data
DP5_2012_5Y <- read_csv(here("data/productDownload_2024-10-19T191549/ACSDP5Y2012.DP05-Data.csv"))

DP5_2012_5Y_edited <- DP5_2012_5Y %>%
  select(-contains("Percent"),
         -contains("Margin"))

head(names(DP5_2012_5Y_edited),5)
# SEX AND AGE
# RACE
# HISPANIC OR LATINO AND RACE
# Total housing units

```



